Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change

### 2023-12-04 ms.qmd
Update intro

1. Phenotypic plasticity and evolutionary adatation to novel environments
2. Capacity for plasticity and adaptation differ between aqautic and terrestrial ectotherms
3. Climate change is expected to impact on the capacity for plasticity and adaptation
4. We test whether climate variability and predictability explain a populations acclimation capacity and affect changes in trait variance.

Climate change is expected to result in warmer and more variable thermal environments for many organisms globally [@Suarez2023]. Greater thermal variability is expected to result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered critical for population resilience to human-induced climate change [@Seebacher2015-wi]. Nonetheless, the extent to which plasticity versus genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established. Without adaptive genetic or plastic responses high rates of extinction are predicted [@Cahill2012]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable. Despite this theoretical expectation, empirical support for this prediction is scant [], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity). Reversible forms of phenotypic plasticity, such as acclimatization, may not be expected to adhere to such predictions.

@@

Reversibly plastic responses, such as acclimitisation, are expected to provide organisms with resilience to climate change by allowing them to adjust their physiology to match the thermal environment they are exposed to [@Seebacher2015-wi].

>>>>

<!-- 
1. Phenotypic plasticity and evolutionary adatation to novel environments
2. Capacity for plasticity and adaptation differ between aqautic and terrestrial ectotherms
3. Climate change is expected to impact on the capacity for plasticity and adaptation
4. We test whether climate variability and predictability explain a populations acclimation capacity and affect changes in trait variance.
-->

Climate change is expected to result in warmer and more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability is expected to result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered critical for population resilience to human-induced climate change [@Seebacher2015-wi; @Merila2014]. Nonetheless, the extent to which adaptive plasticity versus genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without adaptive genetic or plastic responses high rates of extinction are predicted if organisms cannot track suitable habitats [@Cahill2012; @Chevin2017]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable [@Chevin2017]. Despite this theoretical expectation, empirical support for this prediction is scant [], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity). Reversible forms of phenotypic plasticity, such as acclimatization, may not be expected to adhere to such predictions.

@@

Studies investigating reversibly plastic responses, such as acclimitisation, have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability under more extreme thermal conditions [@Seebacher2015-wi; @Havird2020-mj]. Understanding how variability in physiological rates is affected by climate change has important implications for understanding a populations capacity for physiological traits to evolve because lack of physiological variation is expected to limit responses to selection (i.e., the opportunity for selection) [@Sanderson2023]. Change is physiological variability can also inform on the extent to which populations are able to buffer against environmental change [@Sanderson2023]. For example, if populations are able to buffer against environmental change by increasing physiological variability, this may be a mechanism that allows populations to persist in the face of climate change.
